\Ace: Growing A Statically-Typed Language Inside a Python

### 2014-03-24 ace-oopsla14/ace-oopsla14.tex
OOPSLA14 progress

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce Ace, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, Ace is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in Ace can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems. This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the need to resolve link-time ambiguities.

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular type system, the Ace compiler inverts control over type assignment and translation to type-level functions according to a type-directed protocol that does not lead to ambiguities when type systems are combined. We show that this protocol is flexible enough to admit library-based implementations of types  drawn from functional, object-oriented, parallel, low-level and domain-specific languages. We also show how this permits the construction of safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace. We also give a simplified calculus that describes how a type-level language can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts (\emph{providers}) designing potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system. In these situations, abstraction providers often develop a new language or language dialect. Unfortunately, this \emph{language-oriented approach} \cite{language-oriented} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

@@

Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{sugarj} must overcome the same extrinsic issues as a language like HJ. Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss. Too little control, on the other hand, leaves it difficult to implement interesting abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both these extrinsic and intrinsic factors. To address the ``chicken-and-egg'' problem, Ace is implemented within Python, entirely as a library  \cite{Politz:2013:PFM:2509136.2509536,python}. The top-level of an Ace file can be thought of as a \emph{compilation script} written in Python and, as we will discuss, Python serves as Ace's type-level language.  Ace and Python share a common syntax and package system, so users of Ace can leverage established tools and infrastructure directly. %The semantics of  define run-time behavior.

Functions marked by a decorator as under the control of Ace are \emph{statically typechecked}. This makes it possible to rule out many potential issues ahead of execution and thus avoid the need for costly dynamic checks. % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
However, there is \emph{no particular type system built into Ace}. Instead, a type is any  {compile-time object} implementing the \verb|ace.Type| interface. We call these \emph{active types}. When statically assigning a type to a compound form, e.g. \verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g. to the type of \verb|e|, to determine how to assign a type to the expression as a whole. As we will show, this {type-directed protocol}, which we call \emph{active type assignment}, permits the expression of a rich variety of type systems as libraries. Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (extensions have exclusive control over non-overlapping sets of expressions by construction). Base cases, e.g.  variables and literals, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.

@@

The dynamic behavior of an Ace function is determined by translation to a \emph{target language}. We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g. GPUs). Active types and bases control translation by the same type-directed protocol that governs type assignment. We call this phase \emph{active translation}. A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|. 
%One base or type can support multiple targets. 
When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{TIL}) into the language.

Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python scripts. Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions. This occurs both between Ace functions and when calling Ace functions from Python. In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs. This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (common in, for example, scientific computing) that alternate between Python for orchestration and productivity-oriented tasks (e.g. exploratory plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example that simultaneously uses dynamically typed terms (which can be thought of as having the static type \verb|dyn|) alongside statically typed immutable records and a simple prototypic object system, all implemented orthogonally as active types. In Sec. \ref{att}, we detail how active type assignment and translation delegate control, at compile-time, to active types and bases. In Sec. \ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety is maintained. In Sec. \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.  In Sec.  \ref{related}, we compare Ace to related work. We conclude in Sec. \ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work. The appendix contains details omitted from the main body and additional examples.

@@

%
Listing \ref{example} shows an example of an Ace compilation script. As promised, compilation scripts are written directly in Python. Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy). This choice pays immediate dividends on the first five lines: Ace's package system is Python's package system, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing libraries that use Ace. Note that all of the type systems that we discuss are implemented entirely as  libraries.

@@

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} \cite{tlc} and Python is its type-level language (discussed further in Sec. \ref{theory}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will see how they are implemented in Sec. \ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g. \verb|a[min:max]|). Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 8-11, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). Our appendix specifies the typechecking rules, based on \cite{nfulton-thesis}. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and numbers.
\item Prototypic inheritance is a feature of some dynamically-typed languages (most notably, Self \cite{self} and Javascript). On line 12, we consider a simple statically-typed variant. The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores. In our example, we will perform multiple transfers differing only in amount.
\end{enumerate}
\subsection{Functions}
Functions implementing run-time behavior are distinguished by a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below). These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators. We will see how, with a bit of cleverness, Python's syntax can support a variety of static semantics.
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing1.py`
Hello, compile-time world!
[acec] _listing1.py successfully generated.
$ `python _listing1.py`
Hello, run-time world!
Transferring 5.50 to Annie Ace.
Transferring 15.00 to Annie Ace.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}. Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
\label{external-compilation}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
import examples.num.runtime as __ace_0

@@

On lines 14-18, we see the function \verb|log_transfer|. As with a Python function, we write a documentation string like the one shown specifying, informally, the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating simply that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}

Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{If the right-hand side of every annotation is not a type, it will be an error. In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword to skip the line.} % (and following conventions for documentation strings, which also describe ``types'' at the end). % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
 {In version 3.0 of Python, syntax for annotating arguments with arbitrary metadata using \verb|:| was introduced,  intended for use by future tools (further suggesting that users of dynamically typed languages see potential in a static typing discipline)  \cite{PEPwhatever}. Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here to satisfy our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (we will consider cross-compilation in Sec. \ref{targets}, however).

\subsection{Bidirectional Typechecking of Introductory Forms}
On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec. \ref{compilation}). After printing a run-time greeting, we construct a value of type \verb|C| on lines 24-26. Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using syntax for a dictionary literal. We must specify which type the literal should be {analyzed} against by giving a \emph{literal ascription}  \verb|(C)|. This again repurposes existing  syntax: here function application. When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type  (we will see precisely how it is checked shortly). We see the same form used with tuple literals on lines 27-28 to introduce terms of type \verb|Transfer|.

%\subsection{Bidirectional Typechecking}
The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but we do not include an ascription because the definition of \verb|C| determines which types they should be analyzed against. As we will discuss shortly, Ace has a \emph{bidirectional type system} distinguishing locations where a type must be \emph{synthesized} from an expression (e.g. at the top-level) and locations where the expression must be \emph{analyzed} against a known type \cite{LovasPfenning08}. Literal forms can only be analyzed against a type. For example, if they are used in a synthetic context, the base determines a ``default ascription''. If we had not specified an ascription on line 26, the literal would need to synthesize a type, so the base, \verb|py|, would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys are not bound.

An ascription can also be a {type constructor}, rather than a type. For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[int])| assuming the default ascription for integer literals is \verb|int|. The type constructor synthesizes an index based on the types synthesized by the subexpressions. If we wanted a matrix of \verb|float|s, then we would have to write either \verb|[1, 2] (matrix[float])| or \verb|[1(float), 2(float)] (matrix)|. We will see how this works in greater technical detail shortly.

\subsection{External Compilation}\label{compilation} 
To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python. We will first describe the former, then return to interactive compilation in Sec. \ref{targets} with a different example. Listing \ref{external-compilation} shows how to use the \verb|acec| compiler to typecheck and translate \texttt{listing\ref{example}.py}, producing a file \verb|_listing1.py| which can then be executed by the standard Python interpreter. Note that the \textbf{\texttt{print}} statement at the top level of the compilation script was indeed evaluated during compilation. These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case unless requested).

The \verb|acec| compiler (a simple Python script) operates in two stages:
\begin{enumerate}
\item Executes the provided compilation script (\verb|listing1.py|)
\item For any top-level bindings that are Ace functions in the final environment (with dynamic type \verb|ace.ConcreteFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec. \ref{att}. If no type errors are discovered, the translations are extracted and emitted as files (their extensions are determined by the targets in use, discussed in Sec. \ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|). If a type error is discovered, no file is emitted and the error message is displayed on the console.
\end{enumerate}

>>>>

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce Ace, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, Ace is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in Ace can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems. This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the possibility of link-time ambiguities or safety issues.

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular type system, Ace has an extensible bidirectional type system that inverts control over typing and translation to type-level functions according to a protocol that cannot lead to ambiguities and type safety issues when type systems are combined. We give a simplified calculus that describes the type theoretic foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages. We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

@@

Unfortunately, researchers and domain experts aiming to provide potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system. In these situations, abstraction providers often develop a new language or language dialect. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

@@

Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} must overcome the same extrinsic issues as a language like HJ. Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss. Too little control, on the other hand, leaves it difficult to implement interesting abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both extrinsic and intrinsic factors. To address the ``chicken-and-egg'' problem, Ace is implemented within Python \cite{Politz:2013:PFM:2509136.2509536,python}, entirely as a library. The top-level of an Ace file is a \emph{compilation script} written in Python and Python serves as Ace's \emph{type-level language}, as we will discuss.  Ace and Python share a common syntax and package system, so users of Ace can directly leverage established tools, infrastructure and coding standards. %The semantics of  define run-time behavior.

Functions marked by a decorator as under the control of Ace are, unlike Python functions, \emph{statically} typechecked. This makes it possible to rule out many potential issues ahead of execution, and thus  avoid  costly dynamic checks. % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
However, there is \emph{no particular type system built into Ace}. Instead, a type is any  object implementing the \verb|ace.Type| interface constructed by the compilation script. We call these \emph{active types}. When statically assigning a type to a compound form, e.g. \verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g. to the type of \verb|e|, to determine how to assign a type to the expression as a whole. As we will show, this {type-directed protocol}, \emph{active typing}, permits the expression of a rich variety of type systems as libraries. Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (types control non-overlapping sets of expressions by construction). Base cases, e.g.  variables and certain statement forms, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.

@@

The dynamic behavior of an Ace function is determined by translation to a \emph{target language}. We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g. GPUs). Active types and bases control translation by the same type-directed protocol that governs type assignment. We call this phase \emph{active translation}. A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|. 
%One base or type can support multiple targets. 
When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{tarditi+:til-OLD}) into the language.

Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python. Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions. This occurs both between Ace functions and when calling Ace functions from Python. In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs. This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (common in, for example, scientific computing) that alternate between Python for orchestration and productivity-oriented tasks (e.g. exploratory plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example that uses several interesting Ace type libraries. In Sec. \ref{att}, we detail how active typing and translation delegate control, at compile-time, to active types and bases, and discuss how this represents a form of extensibility that side-steps the expression problem. In Sec. \ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety can be maintained. In Sec. \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.  In Sec.  \ref{related}, we compare Ace to related work. We conclude in Sec. \ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work. The appendix contains details omitted from the main body and several additional examples.

@@

%
Listing \ref{example} shows an example of an Ace compilation script. As promised, compilation scripts are written directly in Python. Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy). This choice pays immediate dividends on the first five lines: Ace's import mechanism is Python's import mechanism, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing libraries that use Ace. Note that all of the type systems that we discuss are implemented entirely as  libraries.

@@

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec. \ref{theory} and Sec. \ref{related}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will see how they are implemented in Sec. \ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g. \verb|a[min:max]|). Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 8-11, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). Our appendix specifies the typechecking rules, based on \cite{fulton-thesis}. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records, numbers and strings.
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line 12, we consider a simple \emph{prototypic object type} \cite{Lie86}. The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
\end{enumerate}
\subsection{Functions}
Functions implementing run-time behavior (\emph{typed functions}) are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below). These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators. We will see how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's. Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for typed functions \cite{python}. 
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing1.py`
Hello, compile-time world!
[acec] _listing1.py successfully generated.
$ `python _listing1.py`
Hello, run-time world!
Transferring 5.50 to Annie Ace.
Transferring 15.00 to Annie Ace.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}. Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
\label{external-compilation}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
import examples.num.runtime as __ace_0

@@

On lines 14-18, we see the function \verb|log_transfer|. As in Python, we write a documentation string  that informally specifies the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}

Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end). % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
 {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{PEP}:
% \vspace{-5px}
\begin{lstlisting}[numbers=none]
def log_transfer(t : Transfer): ...
\end{lstlisting}
\vspace{-5px}
These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries. Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (though we consider cross-compilation in Sec. \ref{targets}).

\subsection{Bidirectional Typechecking of Introductory Forms}
On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec. \ref{compilation}). After printing a run-time greeting, we construct a \emph{value} of type \verb|C| on lines 24-26. Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals. We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|. This again repurposes existing  syntax: here function application. When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers and keywords like \verb|True| and \verb|None|) and the argument is a type, we consider it instead an \emph{ascribed introductory form} of that type. We see the same form used with Python's tuple syntax on lines 27-28 to introduce terms of type \verb|Transfer| by providing the fore and prototype.

%\subsection{Bidirectional Typechecking}
The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but do not require an ascription. This is because the definition of the type \verb|C| already determines which types they should be analyzed. As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g. to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g. as an argument to a function with known type) \cite{Lovas08abidirectional}. Literal forms can only be analyzed against a type. If used in a synthetic location, the base must specify a ``default ascription''. For example, if we had not specified the literal ascription on line 26, the base would cause it to be analyzed against a type for Python dictionaries, which would lead to a type error because the variables used as keys are not bound in the surrounding context.

An ascription can also be a {type constructor}, rather than a type. For example, we might write \verb|[1, 2] (matrix)| rather than \verb|[1, 2] (matrix[i32])| assuming the default ascription for integer literals is \verb|i32|. The type constructor synthesizes an index based on the types synthesized by the subexpressions. If we wanted a matrix of \verb|f32|s, then we would have to write either \verb|[1, 2] (matrix[f32])| or \verb|[1(f32), 2(f32)] (matrix)|, or use a base with a different default ascription. We will see how this works in greater technical detail shortly.

\subsection{External Compilation}\label{compilation} 
To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively from within Python. We will here describe the former, then return to interactive compilation in Sec. \ref{targets} with a different example.

Listing \ref{external-compilation} shows how to invoke the \verb|acec| compiler at the shell to typecheck and translate \texttt{listing\ref{example}.py}, resulting in a file named \verb|_listing1.py|. This is then sent to the  Python interpreter to produce the desired run-time behavior. Note that the {\texttt{print}} statement on line 32 of the compilation script was  evaluated during compilation only. These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated in this case).

The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
\begin{enumerate}
\item Executes the provided compilation script (\verb|listing1.py|)
\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates the active type-checking and translation process described in Sec. \ref{att}. If no type errors are discovered, the translations are extracted and emitted (the file extension(s) are determined by the target(s) in use, discussed in Sec. \ref{targets}; here our target is  the default target associated with the base \verb|examples.py.py|, which emits Python 2.6 files). If a type error is discovered, no file is emitted and the error is displayed on the console.
\end{enumerate}
